Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-09-02 paper/paper.tex
latest paper

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
{\bf We argue that the results from the multiphase model 
provide a correct interpretation of the data.
In that case the clump velocity dispersion implies a dynamical mass of
$2\times 10^{9}$ M$_{\odot}$,  ten times its baryonic mass.}
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial \lya emitting galaxies, opening
{\bf the opportunity to constrain models of star formation and
  feedback in the first generation of galaxies.}
\end{abstract}

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya emission with puzzling 
features: the line is highly symmetric, single peaked and broad 
\citep{Thuan97}.
This kind of profile also raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya radiation could emerge as a low surface
brightness glow, which may be connected to \lya halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep[see the review by][and references therein]{2014PASA...31...40D}.

Attempts to explain the atypical \lya features in \tol\ with
conventional models based on a expanding shell have not been successful
so far \citep{mashesse03,2015A&A...578A...7V}. 
This justifies a new study using two new physically motivated
situations: a multiphase outflowing interstellar medium and bulk rotation for a
galaxy without outflows.

The main motivation for the outflowing multiphase model \citep[as presented
in][]{Gronke2016} is that some dwarf galaxies are expected to
present outflows.  
Observationally, outflows have been detected in few local dwarf galaxies
\citep{1998ApJ...506..222M,2005MNRAS.358.1453O}. 
Besides, clumpy multiphase outflows are capable to explain \lya features around
star-forming galaxies
\citep{2010ApJ...717..289S,2012MNRAS.424.1672D}.
In addition, due to the cooling properties of gas, multiphase media are expected
in a range of astrophysical systems
\citep[][]{1977ApJ...218..148M}.

The motivation for the rotation without outflows 
\citep[as presented in][]{GaravitoCamargo2014} is that dwarf galaxies show coherent
rotation features  \citep{2009A&A...493..871S} and it is expected that
some of them have high neutral gas contents with long quiescent phases
without high gas outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.

{\bf Furthermore, observations of other Compact Dwarf Galaxies (CDGs) also show 
that the kinematics range from pure rotation and low velocity
dispersion to high velocity dispersion without rotation
\citep{2015A&A...577A..21C,2017A&A...600A.125C}. }

The models we use correspond to simplified geometrical configurations.
This allows us to perform a deep exploration of parameter space and
gain some physical insight into \tol's kinematic properties.
In this paper we show, for the first time in the literature, that \tol's
\lya profile can be explicitly modeled by either of these two
models. 
{\bf The hybrid case of a rotating multiphase configuration
cannot be simulated with existing numerical implementations and 
is the subject of ongoing research (Remolina-Gutierrez et al. in
prep.) }

In the next section we review the observational characteristics of
\tol, then we summarize the main features in the multiphase and
rotation models to explain how we fit their free parameters 
to the \tol's \lya line shape.
We use the results to interpret them in terms of the galaxy's
dynamical mass and argue why the multiphase model should be preferred
over the rotation model.
We close by discussing the additional observational information needed
to clarify the kinematic nature of \tol.

@@

\begin{table}
\begin{center}
\begin{tabular}{lr}\hline
$\alpha$(2000)& 12h17min17.1s\\
$\delta$(2000)& -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
Redshift & $0.026\pm0.001$ \\
2D half-light radius  & $1.5\pm 0.1$ kpc\\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of \tol\ 
  \citep{Thuan97}.
The 2D half-light radius is
computed from the results reported by \citet{2001AJ....121..169F}.
\label{obstable}}
\end{table}

@@

Qualitatively this result can be understood as follows.
Due to the large fraction of \lya photons being emitted within the
moving clumps ($P_{\rm cl}\sim 1$) the `intrinsic' profile follows 
the clump kinematics. In other words, the width of the intrinsic line
is set by $\sigma_{\mathrm{cl}}$ and its median offset mainly by
$v_{\infty,{\rm cl}}$ and $\beta_{\rm cl}$, {\bf where $\beta_{\rm cl}$ is
the exponent defining the steepness of the radial velocity profile. }
Furthermore, the relatively low mean number of clumps per line of
sight, $0.8<f_{\mathrm{cl}}<8$, combined with the high velocity dispersion
of the clumps implies existence of low-density inter-clump 
regions where \lya photons can freely propagate.
This gives result to an emergent spectrum close to the intrinsic one, explaining the high
flux at the line's center.

@@

However, we cannot discard that another region of parameter space
could also yield a good fit to the observed line-profile.
This might be interesting to explore in the
future with, e.g., a higher sampling of the parameter space once new 
observations  clarify the kinematic nature of \tol. 
In Section \S \ref{sec:conclusions} we describe  in more detail what
kind of observations could be performed in the near future to improve
our understanding of \tol's kinematics.

@@

\tol\ presents an atypical \lya spectrum. 
Its flux at the line's center is high compared to other LAEs at low
redshift and the broad, highly symmetrical peak is virtually absent from other
LAEs at low and high redshift
\citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. 
Simple shell models
fail to reproduce such a spectrum as reported by \cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models: multiphase ISM and solid
body rotation.

{\bf Which model could be closer to the actual kinematic conditions in \tol?  
Recent Integral Field Unit (IFU) observations of other CDGs seem to
favor the multiphase model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial smapling was $0.67^{\prime\prime}$ and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine BCGs to produce two-dimensional maps of the
continuum and promiment emission lines to finally compute velocitity
field maps using the H$\alpha$ emission line. 
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120$ \kms) to highly irregular. 
The typical velocity dispersion values are in the range $10-50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms. }

{\bf These results disfavor the high rotational velocity of $V_{m}\sim350$
\kms\  that we find in the pure rotation model. 
On the other hand the results from the multiphase model yield a
velocity dispersion $\sigma=$\sigmaclump, quite close to observed
values in other CDG, favoring the multiphase model.}
In this case we can  estimate a value for the dynamical
mass using the constraints for the velocity dispersion, $\sigma$,  in
a spherical system localized in a region of size $r$

@@

Assuming that the \lya emission is entirely powered by star formation 
we use the 3D half-light radius $r_s=2.25$ kpc as the typical size
for the HI region, in the multiphase model the clump velocity dispersion
$\sigma_{\rm{cl}}=$\sigmaclump\  corresponds to a dynamical mass of
$M_{\rm dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$, which is ten
times the estimated baryonic mass in \tol.
This would mean that \tol\ is dark matter dominated.

Following the methodology of \citet{2011ApJ...726..108T} we estimate
that a dark matter halo with a virial mass $\sim 6\times 10^{11}$
M$_{\odot}$ {\bf and a virial radius $220$ kpc } could explain this dynamical mass.
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.  
{\bf This estimate should be considered an upper bound as lower values
  could be achieved if one considers instead a cored profile.}
Figure \ref{fig:mass} shows the enclosed mass as a function of radius
for different dark matter halos together with the \tol's dynamical
mass estimate. 
{\bf This estimate and the validity of its assumptions can only be
  verified by detailed mass modelling once \tol's full rotation curve
  becomes available.}

{\bf New IFU observations are needed to confirm \tol's kinematics.
This could be done with the Multi Unit Spectroscopic Explorer (MUSE) 
\citep{2014Msngr.157...13B}, the Gemini Multi-Object Spectrographs (GMOS) 
\citep{2004PASP..116..425H} or the Fibre Large Array Multi Element
Spectrograph (FLAMES) \citep{2002Msngr.110....1P} as they have the spatial
resolution and field of view required to perform the kind of study
presented by \citet{2015A&A...577A..21C} on CDGs or \citet{Herenz16}
on nearby \lya\ emitting galaxies.}

@@

All in all, the mere existence of a strong LAE galaxy with a broad,
highly symmetric line is interesting. 
\tol's line shape is different to others and seems to reveal an
unexpected kinematic structure.
A confirmation of our results by new observations would support the multiphase model
as an element to be included in the study of primordial \lya emitting
galaxies.

>>>>

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
{\bf We argue that the results from the multiphase model 
provide a correct interpretation of the data.
In that case the clump velocity dispersion implies a dynamical mass of
$2\times 10^{9}$ M$_{\odot}$,  ten times its baryonic mass.}
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial galaxies, opening
{\bf the opportunity to use the models and methods presented in this
  paper to constrain the physics of star formation and feedback in the
  early generation of \lya emitting galaxies.} 
\end{abstract}

@@

\tol\ is a compact dwarf galaxy (CDG) that presents a
strong \lya emission with puzzling 
features: the line is highly symmetric, single peaked and broad 
\citep{Thuan97}.
This kind of profile also raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya radiation could emerge as a low surface
brightness glow, which may be connected to \lya halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep[see the review by][and references therein]{2014PASA...31...40D}.

Attempts to explain the atypical \lya features in \tol\ with
conventional models based on a expanding shell have not been successful
so far \citep{mashesse03,2015A&A...578A...7V}.

{\bf Motivated by observations of other CDGs
  that show gas kinematics ranging from pure rotation and low velocity
dispersion to high velocity dispersion without rotation
\citep{2015A&A...577A..21C,2017A&A...600A.125C} we perform here a new
study to explain \tol's \lya emission features under two physical
conditions for the interstellar medium: multiphase outflows and pure rotation.}

Additional motivation for the outflowing multiphase model \citep[as presented
in][]{Gronke2016} is that some dwarf galaxies are expected to
present outflows.  
Observationally, outflows have been detected in few local dwarf galaxies
\citep{1998ApJ...506..222M,2005MNRAS.358.1453O}. 
Besides, clumpy multiphase outflows are capable to explain \lya features around
star-forming galaxies
\citep{2010ApJ...717..289S,2012MNRAS.424.1672D}.
In addition, due to the cooling properties of gas, multiphase media are expected
in a range of astrophysical systems
\citep[][]{1977ApJ...218..148M}.

Further  motivation for the model of pure rotation without outflows 
\citep[as presented in][]{GaravitoCamargo2014} is that dwarf galaxies show coherent
rotation features  \citep{2009A&A...493..871S} and it is expected that
some of them have high neutral gas contents with long quiescent phases
without high gas outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.

The models we use correspond to simplified geometrical configurations.
This allows us to perform a deep exploration of parameter space and
gain some physical insight into \tol's kinematic properties.
In this paper we show, for the first time in the literature, that \tol's
\lya profile can be explicitly modeled by either of these two
models.

In the next section we review the observational characteristics of
\tol, then we summarize the main features in the multiphase and
rotation models to explain how we fit their free parameters 
to the \tol's \lya line shape.
We use the results to interpret them in terms of the galaxy's
dynamical mass and argue why the multiphase model should be preferred
over the rotation model.

@@

\begin{table}
\begin{center}
\begin{tabular}{lr}\hline
$\alpha$(2000)& 12h17min17.1s\\
$\delta$(2000)& -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
Redshift & $0.026\pm0.001$ \\
2D half-light radius  & $1.5\pm 0.1$ kpc\\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of \tol\ 
  \citep{Thuan97}.
The 2D half-light radius is computed from the results reported by
\citet{2003A&A...410..481N}.  
\label{obstable}}
\end{table}

@@

Qualitatively this result can be understood as follows.
Due to the large fraction of \lya photons being emitted within the
moving clumps ($P_{\rm cl}\sim 1$) the `intrinsic' profile closely follows 
the clump kinematics. 
In other words, the width of the intrinsic line
is set by $\sigma_{\mathrm{cl}}$ and its median offset mainly by
$v_{\infty,{\rm cl}}$ and $\beta_{\rm cl}$, {\bf the exponent defining
  the steepness of the radial velocity profile. } 
Furthermore, the relatively low mean number of clumps per line of
sight, $0.8<f_{\mathrm{cl}}<8$, combined with the high velocity dispersion
of the clumps implies existence of low-density inter-clump 
regions where \lya photons can freely propagate.
This gives result to an emergent spectrum close to the intrinsic one, explaining the high
flux at the line's center.

@@

However, we cannot discard that another region of parameter space
could also yield a good fit to the observed line-profile.
This might be interesting to explore in the
future with, e.g., a higher sampling of the parameter space once new 
observations  yield more details on \tol's kinematic structure.

@@

{\bf \tol\ presents a \lya emission line with puzzling fetures.}
Its flux at the line's center is high compared to other LAEs at low
redshift and the broad, highly symmetrical peak is virtually absent from other
LAEs at low and high redshift
\citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. 
Simple shell models
fail to reproduce such a spectrum as reported by \cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models: multiphase ISM and solid
body rotation.

{\bf Which model could be closer to the actual kinematic conditions in \tol?  
Integral Field Unit (IFU) observations of other CDGs seem to
favor the multiphase model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial smapling was $0.67^{\prime\prime}$ and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine BCGs to produce two-dimensional maps of the
continuum and promiment emission lines to finally compute velocitity
field maps using the H$\alpha$ emission line. 
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120$ \kms) to highly irregular. 
The typical velocity dispersion values are in the range $10-50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms. }

{\bf These results disfavor the high rotational velocity of $V_{m}\sim350$
\kms\  that we find in the pure rotation model. 
On the other hand the results from the multiphase model yield a
velocity dispersion $\sigma_{\rm cl}=$\sigmaclump, consistent with observed
values in other CDG, favoring the multiphase model.}
In this case we can  estimate a value for the dynamical
mass using the constraints for the velocity dispersion, $\sigma$,  in
a spherical system localized in a region of size $r$

@@

Assuming that the \lya emission is entirely powered by star formation 
we use the 3D half-light radius $r_s=2.25$ kpc as the typical size
for the HI region and that  $\sigma=\sigma_{\rm{cl}}=$\sigmaclump, we
obtain a dynamical mass of  $M_{\rm dyn}=2.31\pm0.04 \times 10^{9}$
$M_{\odot}$, which is ten times the estimated baryonic mass in
\tol.

A larger dynamical mass estimate over its baryonic mass is a hint that
\tol\ is dark matter dominated.  
Following the methodology of \citet{2011ApJ...726..108T} we estimate
that a dark matter halo with a virial mass $\sim 6\times 10^{11}$
M$_{\odot}$ {\bf and a virial radius $220$ kpc } could explain this dynamical mass.
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.  
Figure \ref{fig:mass} shows the enclosed mass as a function of radius
for different dark matter halos together with the \tol's dynamical
mass estimate. 
{\bf This value should be considered an upper bound as lower values
  could be achieved if one considers instead a cored profile.}
{\bf More realistic mass estimates can only achieved by detailed mass
  modelling once \tol's full rotation curve   becomes available.}

{\bf New IFU observations are needed to confirm \tol's kinematics.
This could be done with the Multi Unit Spectroscopic Explorer (MUSE) 
\citep{2014Msngr.157...13B}, the Gemini Multi-Object Spectrographs (GMOS) 
\citep{2004PASP..116..425H} or the Fibre Large Array Multi Element
Spectrograph (FLAMES) \citep{2002Msngr.110....1P} as they have the spatial
resolution, spectral coverage and field of view required to produce
H$\alpha$ velocity maps to perform the kind of study presented by
\citet{2015A&A...577A..21C} on CDGs or \citet{Herenz16} on nearby
\lya\ emitting galaxies.}

@@

All in all, the mere existence of a strong LAE galaxy with a broad,
highly symmetric line is interesting. 
\tol's line shape is different to others and seems to reveal an
unexpected kinematic structure.
A confirmation of our results by new observations would support the multiphase model
as an element to be included in the study of \lya emitting
galaxies, {\bf consolidating the possibility to use the \lya line to
  constrain different models of star formation and feedback in the
  first generation of galaxies. }
